Tropical mathematics for the Lambda-calculus

### 2023-07-27 CSL24/section1.tex
Commit notturno

On the one hand one there is the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function, it makes more sense to ask whether they compute functions which do not differ \emph{too much} by controlling the propagation of an error during the computation.
This has motivated the study of denotational frameworks in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{Reed2010} and coinductive methods \cite{Bonchi2018}, and was recently extended to account for the full $\lambda$-calculus \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.
%As already mentioned, in many situations (e.g.~when dealing with computationally difficult problems, or in differential privacy \cite{Alvim2011, Reed2010}), one %does not look for algorithms to compute a function \emph{exactly}, but rather to approximate it (in an efficient way) within some error bound. In other common situations (e.g.~in differential privacy \cite{Alvim2011, Reed2010}) one needs to verify that an algorithm is not \emph{too sensitive} to errors, that is, a small error in the input will produce a comparably small error in the output. 
%In all these cases, it is common to consider forms of denotational semantics in which types are endowed with a \emph{behavioral metric}, that is, a metric on programs which account for differences in behavior. 
A fundamental insight coming from this line of work is that \emph{affine} programs, i.e.~programs that may use their input at most once, correspond to \emph{non-expansive} (or $1$-Lipschitz) maps, that is, to functions $f$ for which the distances
$d(f(x),f(y))$ produced in output are bounded by the distances $d(x,y)$ in input. 
This is due to the fact that the category $\Met$ of pseudo-metric spaces and non-expansive maps provides a model of the %\emph{linear} simply typed $\lambda$-calculus, being a \emph{symmetric monoidal closed} category, and in fact it also models 
\emph{affine} $\lambda$-calculus (it is symmetric monoidal closed, cartesian and monoidal units coincide). 
The metric approach is not restricted to affine programs, but can be extended to programs with \emph{bounded} duplications \cite{Reed2010, Gaboardi2017}: the idea is that a program duplicating its input $K$ times will give rise to a $K$-\emph{Lipschitz} map.
For instance, the higher-order program $M=\lambda f.\lambda x.f(f(x))$, which duplicates the function input $f$, yields a $2$-Lipschitz map between the metric space $\BB R\multimap \BB R$ of non-expansive real functions and itself. %: if $f,g$ are two non-expansive maps differing by at most $\epsilon$ (i.e.~for which $|f(x)-g(x)|\leq \epsilon$ holds for all $x\in \BB R$), then the application of $M$ to $f$ and $g$ will produce two maps differing by at most $2\epsilon$. 
By observing that a $r$-Lipschitz map between metric spaces $X$ and $Y$ is the same as a non-expansive map between the \emph{re-scaled} space $rX:=(X,%d_{r}(x,y)=
rd(x,y))$ and $Y$, the program $M$ above 
can thus be interpreted as a non-expansive map from $2(\BB R\multimap \BB R)$ to $\BB R\multimap \BB R$.

On the other hand, there is the approach based on \emph{differential} \cite{difflambda} or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected to the so-called \emph{relational semantics} \cite{Manzo2012, Manzo2013, dill} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}. This family of approaches have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{Accattoli2021}, and have also been shown suitable for the probabilistic setting \cite{Manzo2013, Breuvart2018, PistoneLICS2022}. 
%A different family of approaches to linearity and duplication arises from the study of the \emph{differential $\lambda$-calculus} \cite{difflambda} (and differential linear logic \cite{dill}) and its categorical models. 
The key ingredient is a \emph{differential operator} $\Der$,  added to the usual syntax of the $\lambda$-calculus. The intuition is that, given $M$ of type $A\to B$ and $N$ of type $A$, the program $\Diff{M}{N}$, still of type $A\to B$, corresponds to the \emph{linear application} of $M$ to $N$: this means that $N$ is passed to $M$ so that the latter may use it exactly once (this is why $\Diff{M}{N}$ still has type $A\to B$, since $M$ might need \emph{other} copies of an input of type $A$).

>>>>

On the one hand one there is the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function, it makes more sense to ask whether they compute functions which do not differ \emph{too much} by controlling the propagation of an error during the computation.
This has motivated the study of denotational frameworks in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{Reed2010} and coinductive methods \cite{Bonchi2018}, and was recently extended to account for the full $\lambda$-calculus \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.
%As already mentioned, in many situations (e.g.~when dealing with computationally difficult problems, or in differential privacy \cite{Alvim2011, Reed2010}), one %does not look for algorithms to compute a function \emph{exactly}, but rather to approximate it (in an efficient way) within some error bound. In other common situations (e.g.~in differential privacy \cite{Alvim2011, Reed2010}) one needs to verify that an algorithm is not \emph{too sensitive} to errors, that is, a small error in the input will produce a comparably small error in the output. 
%In all these cases, it is common to consider forms of denotational semantics in which types are endowed with a \emph{behavioral metric}, that is, a metric on programs which account for differences in behavior. 
A fundamental insight coming from this line of work is that \emph{affine} programs, i.e.~programs that may use their input at most once, correspond to \emph{non-expansive} (or $1$-Lipschitz) maps, that is, to functions $f$ for which the distances
$d(f(x),f(y))$ produced in output are bounded by the distances $d(x,y)$ in input. 
This is due to the fact that the category $\Met$ of pseudo-metric spaces and non-expansive maps provides a model of the %\emph{linear} simply typed $\lambda$-calculus, being a \emph{symmetric monoidal closed} category, and in fact it also models 
\emph{affine} $\lambda$-calculus (it is symmetric monoidal closed, cartesian and monoidal units coincide). 
The metric approach is not restricted to affine programs, but can be extended to programs with \emph{bounded} duplications \cite{Reed2010, Gaboardi2017}: the idea is that a program duplicating its input $K$ times will give rise to a $K$-\emph{Lipschitz} map.
For instance, the higher-order program $M=\lambda f.\lambda x.f(f(x))$, which duplicates the function input $f$, yields a $2$-Lipschitz map between the metric space $\BB R\multimap \BB R$ of non-expansive real functions and itself. %: if $f,g$ are two non-expansive maps differing by at most $\epsilon$ (i.e.~for which $|f(x)-g(x)|\leq \epsilon$ holds for all $x\in \BB R$), then the application of $M$ to $f$ and $g$ will produce two maps differing by at most $2\epsilon$. 
By observing that a $r$-Lipschitz map between metric spaces $X$ and $Y$ is the same as a non-expansive map between the \emph{re-scaled} space $rX:=(X,%d_{r}(x,y)=
rd(x,y))$ and $Y$, the program $M$ above 
can thus be interpreted as a non-expansive map from $2(\BB R\multimap \BB R)$ to $\BB R\multimap \BB R$.
These observations have led to the study of linear $\lambda$-calculi with \emph{graded} exponentials, like $\mathsf{Fuzz}$ \cite{Reed2010}, inspired from Girard's Bounded Linear Logic \cite{Girard92tcs}, which have been applied to the study of differential privacy \cite{Gaboardi2013, Gaboardi2017}. The types of such systems are defined by combining linear constructors with a \emph{graded linear exponential comonad} $!_{r}(-)$ \cite{Katsumata2018}.

On the other hand, there is the approach based on \emph{differential} \cite{difflambda} or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected to the so-called \emph{relational semantics} \cite{Manzo2012, Manzo2013, dill} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}. This family of approaches have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{Accattoli2021}, and have also been shown suitable for the probabilistic setting \cite{Manzo2013, Breuvart2018, PistoneLICS2022}. 
%A different family of approaches to linearity and duplication arises from the study of the \emph{differential $\lambda$-calculus} \cite{difflambda} (and differential linear logic \cite{dill}) and its categorical models. 
The key ingredient is a \emph{differential operator} $\Der$,  added to the usual syntax of the $\lambda$-calculus. The intuition is that, given $M$ of type $A\to B$ and $N$ of type $A$, the program $\Diff{M}{N}$, still of type $A\to B$, corresponds to the \emph{linear application} of $M$ to $N$: this means that $N$ is passed to $M$ so that the latter may use it exactly once (this is why $\Diff{M}{N}$ still has type $A\to B$, since $M$ might need \emph{other} copies of an input of type $A$).
Another intriguing similarity between program derivatives and 
actual derivatives is provided by the \emph{Taylor expansion} $\C T$:
%In analysis, a sufficiently regular function can be seen (locally around a point) as its Taylor expansion, i.e.\ a series of polynomials weighted via some rational coefficient.
%Now, a polynomial in $x$ of degree $k$ can be thought of as a function which uses its argument $x$ exactly $k$ times.
in $\STDLC$ (even untyped), we can perform a syntactic Taylor expansion of an ordinary $\lambda$-term via an inductively defined map $\Theta$ giving rise to an infinite series of terms, whose crucial case of the definition is
$
\Theta(MN)=\sum_{n\in\mathbb{N}}\frac{1}{n!}\Der^n\left[\Theta(M),\Theta(N)^n\right]0.
$
As in analysis, it decomposes an application as a series of $k$-linear functions, which can be seen as its approximants.
In other words, unbounded duplications correspond to some sort of limit of bounded, but arbitrarily large, ones.
Since we only consider idempotent sum, the factorial coefficient disappears, and the resulting Taylor map is called the \emph{qualitative} Taylor expansion, indicated by $\Te{\_}$.
